TODO: Funky active learning pun

### 2025-01-09 doc/paper.tex
Finish Introduction and Related works

\section{Introduction}\label{sec:intro}
Complex tasks in supervised learning usually require large annotated datasets. However, the cost and time needed to label samples can be high, especially when expert knowledge in the according field is required.  \\
Active learning aims to reduce the overall labeling cost by selecting the most informative samples for annotation by an oracle and thus enable using machine learning for tasks where annotated data is sparse. \\
The effectiveness of active learning relies heavily on the choice of query strategy, as it determines which samples are selected for annotation and thus directly influences model performance. In pool-based active learning, a model is trained on a small labeled dataset and used to evaluate a large pool of unlabeled data. Informative samples are selected based on a query strategy, labeled by an oracle, and added to the training set. The model is retrained, and the process is repeated until a stopping criterion is met. \\ 
In stream-based active learning, unlabeled data arrives sequentially. Thus, the model decides whether to query the label of an incoming instance using an evaluation metric like the uncertainty of the prediction and checking whether a threshold is reached. If not, the instance is discarded. \\
This report focuses on comparing several pool-based as well as stream-based query strategies. The modular Active Learning framework (modAL)~\cite{danka_modalmodularactivelearning} is used in the following experiments as it simplified the implementation of active learning pipelines, contains multiple already implemented querying strategies and is compatible with models from the Scikit-learn library~\cite{scikit-learn}. \\
Several previous works explore different query methods for different tasks. \\
\cite{schröder_revisitinguncertaintybasedquerystrategies} compares various uncertainty based query strategies in the context of fine-tuning transformer models for text classification, while \cite{zhan_comparativesurveydeepactive} compares several query methods on MNIST and CIFAR. \\
According to \cite{ueno_benchmarkingofquerystrategies} most papers concentrate on two image-based datasets: MNIST~\cite{mnist} and CIFAR~\cite{cifar}, while their paper evaluates strategies on six different datasets, including medical and visual inspection images. They also verify the effectiveness of query strategies independently of the training process, underfitting and randomness due to early-stage training in an experiment where fully trained models are used to select the most informative samples according to various query strategies and to construct a labeled dataset. \\ 
\cite{werner_comparableactivelearning} agrees with their statement that experimental settings are not standardized across papers which makes comparisons of different existing and new query strategies difficult. They propose a benchmark framework evaluating active learning strategies on tabular, image, and text datasets using robust evaluation protocols to mitigate variance and ensure comparability. \\
While \cite{ueno_benchmarkingofquerystrategies} evaluates strategies across several image-based datasets from a wide range of image-domains like the medical and visual inspection images, \cite{werner_comparableactivelearning} focuses on a wide range of data-domains including images, text and vector-based datasets. \\
This report focuses on comparing strategies in the image domain. It compares query strategies for a simple logistic regression model and a more complex CNN to study the effect of model complexity on the optimal query strategy. \\
It is uncertain whether query strategies which work well on simple models generalize well to more complex models like Neural Networks~\cite{schröder_surveyactivelearningtext}. To guarantee real-world applicability, knowledge about this needs to be gained as model complexities vary across different tasks and datasets. Furthermore, the impact of different sizes of the initial annotated dataset is additionally investigated which is something that few studies analyze in combination. The effect of the combination between model complexity and initial annotated dataset size is a gap not researched in previously mentioned papers. \\
By examining different starting sizes, this work simulates various real-world scenarios: (1) starting with no labeled data, (2) having a small pool of labeled data, and (3) working with a well-trained model to identify data points for fine-tuning.

@@

	\item \textbf{Systematic evaluation of active learning query strategies} on MNIST with a logistic regression model and CIFAR-10 using a CNN to analyze their behavior across datasets and model complexities.
	\item \textbf{Investigation of model complexity's influence} on the effectiveness of query strategies to provide knowledge into how shallow and deep models respond differently to active learning methods.
	\item \textbf{Analysis of the initial labeled dataset size's impact} on the performance of query strategies to simulate real-world scenarios such as starting completely from scratch, having a small labeled dataset or fine-tuning a model.

@@

\textcolor{red}{Should I even give a structure of the paper?}

>>>>

\section{Introduction and Related Work}\label{sec:intro}
Supervised learning often requires large, annotated datasets, but labeling data can be costly and time-intensive, especially when expert knowledge is needed. 
Active learning addresses this challenge of reducing the overall labeling costs through the selection of the most informative samples for annotation based on a query strategy. This approach enables machine learning models to perform well in tasks with limited data and high labeling costs. \\
The effectiveness of active learning depends heavily on the chosen query strategy, which determines which samples are selected for annotation. Two main frameworks are commonly used:
\begin{enumerate}
	\item Pool-based active learning: A model is trained on a small labeled dataset and used to evaluate a large pool of unlabeled data. The most informative samples are selected based on a query strategy, labeled by an oracle, and added to the training set. The model is retrained with the updated training set and the process is repeated until a stopping criterion is met.
	\item Stream-based active learning: Here, unlabeled data arrives sequentially and the model decides whether to query the label of an incoming instance using an evaluation metric like the uncertainty of its prediction and checking whether a certain threshold is reached. If not, the instance is discarded.
\end{enumerate}
This report focuses on comparing several query strategies for both pool-based and stream-based active learning. Experiments in this paper leverage the modular Active Learning framework (modAL)~\cite{danka_modalmodularactivelearning}, which simplified the implementation of active learning pipelines, includes multiple pre-implemented query strategies and is compatible with models from the Scikit-learn library~\cite{scikit-learn}. \\
Several previous works have explored and compared different query methods across various domains. \\
\cite{schröder_revisitinguncertaintybasedquerystrategies} compare various uncertainty-based query strategies in the context of fine-tuning transformer models in text classification tasks. \cite{zhan_comparativesurveydeepactive} on the other hand compares several query strategies on the image datasets MNIST and CIFAR. \\
According to \cite{ueno_benchmarkingofquerystrategies} most papers concentrate on two image-based datasets: MNIST~\cite{mnist} and CIFAR~\cite{cifar}, while their paper evaluates strategies on six different datasets, including medical and visual inspection images. They also examine strategies independent of training processes to minimize biases from early-stage randomness or underfitting. This is shown in an experiment where fully trained models are used to select the most informative samples based on several different query strategies to construct a labeled dataset. \\ 
A recurring issue in active learning research is the lack of standardization in experimental setups, as noticed additionally by~\cite{werner_comparableactivelearning}. This inconsistency complicates the comparability of existing and novel query strategies. To address this, they propose a benchmark framework for evaluatting active learning strategies across tabular, image and text datasets. This framework uses robust evaluation protocols to reduce variance and ensure comparability. \\
In contrast, \cite{ueno_benchmarkingofquerystrategies} focus specifically on image-based datasets from diverse domains such as medical or visual inspection images. While~\cite{werner_comparableactivelearning} focuses on a wide range of data-domains including images, text and vector-based datasets, this report focuses on comparing strategies in the image domain. 
It compares query strategies for both models such as simple logistic regression on MNIST and more complex architectures, like a CNN on CIFAR-10 to study the effect of model complexity on the optimal query strategy. \\
It is uncertain whether well working query strategies generalize well from simple to more complex models like Neural Networks~\cite{schröder_surveyactivelearningtext}. Understanding this generalizability is crucial for real-world applications where model complexity varies across tasks and datasets. \\
Furthermore, this report investigates the impact of the initial labeled dataset size, as its impact in combination with model complexity is rarely explored. 
By examining scenarios with varying amounts of initial labeled data, this work simulates various real-world applications: (1) starting with no labeled data, (2) having a small pool of labeled data, and (3) working with a well-trained model to identify data points for fine-tuning.

@@

	\item \textbf{Systematic evaluation of active learning query strategies} across different datasets and model complexities, using MNIST with logistic regression and CIFAR-10 with a CNN.
	\item \textbf{Investigation of the incluence of model complexity} on the effectiveness of query strategies to provide insights into how shallow and deep models respond differently to active learning methods.
	\item \textbf{Analysis of impact of initial labeled dataset size} on the performance of query strategies, simulating possible real-world scenarios such as starting from scratch, having a small labeled dataset or fine-tuning a pre-trained model.

@@

Section~\ref{sec:methods} outlines the used datasets, processing steps and models. Section~\ref{sec:experiments} describes the experimental design, followed by the results and discussion in sections~\ref{sec:results} and~\ref{sec:discussion} respectively. The report concludes with key findings in section~\ref{sec:conclusion}
